---
title: "Prototype Analysis for JIBE project"
author: "Carl Higgs"

output:
    github_document:
        toc: TRUE

---

# Prototype Analysis for JIBE project

This Rmarkdown file provides an example multimodal analysis from origins to destinations to inform modelling in the JIBE project.

It uses r5r, an implementation of Conveyal's R5 routing engine for R.  The R5 software evolved from OpenTripPlanner v1, which was its prototype, and is [recommended](http://docs.opentripplanner.org/en/latest/Version-Comparison/) for use for analytical scenarios such as this one. The r5r library was also recommended as the fastest option in a recent paper, *Higgins, C., Palm, M., DeJohn, A., Xi, L., Vaughan, J., Farber, S., Widener, M., & Miller, E. (2022). Calculating place-based transit accessibility: Methods, tools and algorithmic dependence. Journal of Transport and Land Use, 15(1), 95–116. https://doi.org/10.5198/jtlu.2022.2012*.  The analysis drew upon the r5r [vignette](https://cran.r-project.org/web/packages/r5r/vignettes/intro_to_r5r.html).

# Prerequisites:
The installation of Java 11 and PostgreSQL is required, along with r5r and other packages used in the code example.

The analysis was run using R 4.1.3, within RStudio 1.4

```{r echo=False, include=FALSE}
install.packages('r5r')
install.packages('keyring')
install.packages('RPostgres')
install.packages('progress')

library(keyring)
## Set the following secrets, if not already set
# key_set('database_user')
# key_set('database_password')

```

# Setup
```{r echo=False, include=FALSE}
# Increase memory available
options(java.parameters = "-Xmx4G")

#Load libraries
library(r5r)
library(data.table)
library(parallel) # For parallel processing
library(DBI)      # used to connext to Postgresql using RPostgres
library(progress)

# Set the path to your project directory
data_path <- "C:/Users/E33390/OneDrive - RMIT University/projects/JIBE/multimodal_analysis_example"
list.files(data_path)

# Set up database connection (assumes you have already created the output database)
# e.g. by entering the following in psql:
# CREATE DATABASE multimodal_melbourne_2019;

port <- 5432
host <- 'localhost'
db <- 'multimodal_melbourne_2019'
```
The project directory should contain:
- an OSM excerpt for the region of interest, (http://download.openstreetmap.fr/extracts/oceania/australia/)
- a GTFS.zip file (https://transitfeeds.com/p/ptv/497/20190920/download), 
- and data containing origin and destination locations in EPSG:4326 CRS


# Analysis
```{r}
points <- fread(file.path(data_path, "workEduTrips_longlat.csv"))

origins <- data.table()
origins[,c('id','lon','lat')] <- points[,c('id','ORIGLONG','ORIGLAT')]
origins[,id:=as.character(id)] 

destinations <- data.table()
destinations[,c('id','lon','lat')] <- points[,c('id','DESTLONG','DESTLAT')]
destinations[,id:=as.character(id)] 

# set up
r5r_core <- setup_r5(data_path = data_path, verbose = FALSE)

# set inputs
modes <- c("WALK","BICYCLE","CAR","BUS","RAIL","TRAM")
departure_datetime <- as.POSIXct("16-10-2019 07:45:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# Create empty database tables to hold results
for (mode in modes){
      # Create table to hold results (coordinates not stored; they could be linked if required)
      pg.RPostgres <- dbConnect(RPostgres::Postgres(), 
                            dbname   = db,
                            host     = host,
                            port     = port,
                            user     = key_get('database_user'),
                            password = key_get('database_password'))
      create_table <- paste0(
        "CREATE TABLE IF NOT EXISTS ",
         mode,
        ' (
        "id" integer,
        "option" integer,
        "segment" integer,
        "mode" varchar(8),
        "total_duration" real,
        "segment_duration" real,
        "wait" real,
        "distance" integer)
        ')
      res <- dbSendQuery(pg.RPostgres, create_table)
    
    # Clean up and close database connection
    dbClearResult(res)
}

# Prepare to loop over chunks to mitigate risk of memory error, and slowing associated with this
records <- nrow(origins)
chunk_size <- 100
n_chunks <- ceiling(records / chunk_size)
pb <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed time: :elapsedfull || Estimated time remaining: :eta]",
                       total = n_chunks,
                       complete = "=",   
                       incomplete = "-", 
                       current = ">",    
                       clear = FALSE,    
                       width = 100)   

for(i in 1:n_chunks){  
  pb$tick()
  lower = ((i-1)*chunk_size+1)
  upper = min(records,(i*chunk_size))
  for (mode in modes){
    # calculate OD results
    # For 'walk' mode, walking is unbounded
    # for other modes however, there is a maximum of 1600m walk (about 15 mins)
    # This avoids the observed RAIL restricted scenario of 
    #   - walk 47 mins for train, 
    #   - catch train for 2 mins, 
    #   - then walk further 40 mins
    # To limit trip duration, we allow 8 hours for walking and 4 hours for PT
    dit <- detailed_itineraries(  
                        r5r_core,
                        origins[lower:upper,],
                        destinations[lower:upper,],
                        mode = mode,
                        mode_egress = "WALK",
                        departure_datetime = departure_datetime,
                        max_walk_dist = ifelse(mode=='WALK', Inf, 1600),
                        max_bike_dist = Inf,
                        max_trip_duration =  ifelse(mode=='WALK', 480L, 240L),
                        walk_speed = 3.6,
                        bike_speed = 12,
                        max_rides = 3,
                        max_lts = 2,
                        shortest_path = TRUE,
                        n_threads = Inf,
                        verbose = FALSE,
                        progress = FALSE,
                        drop_geometry = TRUE
                      )
    
    # open Postgres connection
    pg.RPostgres <- dbConnect(RPostgres::Postgres(), 
                          dbname   = db,
                          host     = host,
                          port     = port,
                          user     = key_get('database_user'),
                          password = key_get('database_password'))
    insert_query <- paste("INSERT INTO",mode,"VALUES($1, $2, $3, $4, $5, $6, $7, $8)")
    # insert simulation result to as database row 
    res <- dbSendQuery(pg.RPostgres, 
                       insert_query, 
                       params=unname(as.list(
                          dit[,list(
                                as.integer(fromId),
                                as.integer(option),
                                as.integer(segment),
                                as.character(mode),
                                as.numeric(total_duration),
                                as.numeric(segment_duration),
                                as.numeric(wait),
                                as.integer(distance)
                                )])))
    # clean up and release connection
    dbClearResult(res)
    dbDisconnect(pg.RPostgres)
  }
}

stop_r5(r5r_core)
rJava::.jgc(R.gc = TRUE)
```

Processing the 19,064 records across the 6 separate modes of walking, cycling, car, bus, rail and tram, with allowance for walking egress of up to 1600m and maximum total trip duration of 2 hours, took approximately 32 minutes.  The analysis was run on an x64 computer running Windows 11 Pro for Workstations, with Intel Xeon 3.4Ghz 6-core processor with 32 GB ram.
